---
title: "Project_3"
author: "Robert Sellers & Chirag Vithalani"
date: "March 15, 2016"
output: html_document
---

**Libraries Required**

```{r eval=FALSE}
library(data.table)
library(dplyr)
library(stringr)
library(RMySQL)
```

**Twitter API authentication.** 

**This requires a twitter account, and specific access values per each of the setup_twitter_oauth function parameters.**

***Do not attempt this code without credentials. Please visit https://dev.twitter.com/oauth for further information.*** 

```{r eval=FALSE}
#Warning: Values must be retrieved via twitter account and API / OAuth authentication
############################DO NOT RUN############################
setup_twitter_oauth(consumer_key, consumer_secret, access_token, access_secret)
twitter_results_March_XX<-searchTwitter("#datascience", n=10000)
twitter_results_March_XX<- Map(as.data.frame, twitter_results_March_XX)
twitter_results_March_XX<- rbindlist(twitter_results_March_XX) #uses data.table library
############################DO NOT RUN############################
```

**Or to skip authentication and load the twitter results directly, please run the following instead:**

```{r}
twitter_results_march_16<-read.csv(file="https://raw.githubusercontent.com/RobertSellers/SlackProjects/master/data/twitter_results_March_16.csv", header=TRUE, sep=",")
twitter_results_march_18<-read.csv(file="https://raw.githubusercontent.com/RobertSellers/SlackProjects/master/data/twitter_results_March_18.csv", header=TRUE, sep=",")
twitter_results_march_19<-read.csv(file="https://raw.githubusercontent.com/RobertSellers/SlackProjects/master/data/twitter_results_March_19.csv", header=TRUE, sep=",")
twitter_results_march_20<-read.csv(file="https://raw.githubusercontent.com/RobertSellers/SlackProjects/master/data/twitter_results_March_20.csv", header=TRUE, sep=",")
```

**MySQL Connection. Please ensure that you have run Populate_mySQL.R and P3 SQL DB Create.sql prior to attempting a connection.**

```{r, eval=FALSE}
con <- dbConnect(MySQL(), 
    user = 'root',
    password = 'password', #fill in per user
    host = 'localhost',
    dbname='ds_skills')
#Skills lookup	
lutSkills = dbReadTable(con, "skills")
```

**Lookup Frequencies Function**

```{r}
lookupFrequencies <-function(twitterData,lookupTable) {
  lookupTable<-as.data.frame(lookupTable)
  
  lookupTable$Skill<-paste0("\\<",lookupTable$skill_name,"\\>")
  lookupTable$Skill[20] <- "xxxxxxxxxx" #C++ not working as a keyword
  i<-1
  for(i in 1:nrow(lookupTable)) {
      lookupTable$t_freq[i]<-length(grep(lookupTable$Skill[i], twitterData$text))
      lookupTable$dates[i]<-as.character(as.Date(twitterData[1,6]))
  }
  lookupTable$Skill <- NULL #superfluous
  lookupTable$sc_id <- NULL #superfluous
  lookupTable$skill_name <- NULL #superfluous
  lookupTable$dates <- as.Date(lookupTable$dates)
  return (lookupTable)
}
```

**Data tidying**

```{r, message=FALSE}
#Run the function per raw data result
job1<-lookupFrequencies(twitter_results_march_16,lutSkills)
job2<-lookupFrequencies(twitter_results_march_18,lutSkills)
job3<-lookupFrequencies(twitter_results_march_19,lutSkills)
job4<-lookupFrequencies(twitter_results_march_20,lutSkills)

#combine the data
twitter_freq  <- rbind(job1,job2,job3,job4) 
```

**Write the results to MySQL**

```{r, message=FALSE}
#overwrite/update the table
dbWriteTable(con,value=twitter_freq, 
             name="twitter_freq", 
             append = TRUE, 
             row.name = FALSE,
             field.types=list("int(11)", "int(11)","date")) 

#disconnect from database
dbDisconnect(con)
```