---
title: "Per Capita Metrics as <br> Predictors of Tuberculosis Infection Rates"
author: "James Topor"
date: "May 19, 2016"
output:
  ioslides_presentation:
    widescreen: true
    smaller: true
---

```{r, message=FALSE, warning=FALSE, echo = FALSE}
# Load Packages
library(knitr)
library(RODBC)
library(ggplot2)
suppressMessages(library(rworldmap))
suppressMessages(library(maptools))
library(dplyr)
library(tidyr)
options(stringsAsFactors = FALSE)
```

```{r, echo = FALSE, message = FALSE, warning = FALSE, eval=TRUE}
# establish connection to local SQL server
# NOTE: Be sure to set the server reference appropriately to comport with your own local 
# computing environment. 
con <- odbcConnect("local_server")

# select a database to use - in this case, the 'tb_prediction' database
x <- sqlQuery(con, "use tb_prediction")
rm(x)
```

## Tuberculosis: A Global Problem

__The World Health Organization Says:__

*"Tuberculosis is one of the top 5 causes of death globally among women aged 15-44, and in 2014 alone killed approximately 1.5 million people worldwide."* 

<br> 

__Research Questions:__

- Do countries that have high incidence of TB  share certain __*measurable*__ attributes?

- Do countries that have relatively lower incidence of TB share certain __*measurable*__ attributes that countries with high incidences of TB should perhaps strive to emulate?



## Data Used

__Tuberculosis Infection Counts__

- From Assignment 3: TB Case Counts for 100 countries (1995 - 2013)

__Country Population Counts__

- From Assignment 3: Population counts for 100 countries (1995 - 2013).

__World Bank Data__:

- Life Expectancy at Birth (in years) per Country (1995 - 2013);

- Health Care Expenditure Per Capita per Country (1995 - 2013);

- Gross National Income (GNI) Per Capita per Country (1995 - 2013);

- The percentage of a country's population having access to electricity (2000, 2010, 2012)

__United Nations Data__:

- Average years of Schooling per country (2000, 2005 - 2012)


## Data Acquisition

__All data is downloaded / scraped, transformed / 'tidied', and then stored in MySQL__

| Data                 | Source       | Format      | 
| -------------------- | ------------ | ----------- |
| TB Infection Counts  | MySQL Table  | "Long"      |
| Population Counts    | CSV File     | "Long"      |
| World Bank Data      | CSV Files (4)| "Wide"      |
| UN Data              | Web Scrape   | HTML,"Wide" |

<br>

__*Challenges*:__

- Each data source + R's __rworldmap__ package all use different country naming conventions

- Multiple country name "lookup" tables had to be created by hand in MySQL

- Lack of consistency in the calendar years covered by data sets


## Analysis

- __TB Case Counts vs. TB Infection Rates: Geographical "Hotspots"__

<br> 

- __Trends in TB Case Counts & Infection Rates__

<br> 

- __Per Capita Metrics vs. TB Infection Rates__


## Average Annual TB Case Counts (1995 - 2013)

```{r, echo = FALSE}
# get counts of cases by country, year
tb_df <- sqlQuery(con, "SELECT country, year, SUM(child + adult + elderly) FROM tb
						GROUP BY country, year ORDER BY country, year", stringsAsFactors=F)

# rename third column to meaningful name
colnames(tb_df)[3] <- "cases"

# Ensure country names match those used by R's geomapping tools:
# first, fetch R country names from database lookup table
rmap_df <- sqlQuery(con, "SELECT * FROM rmap_lookup", stringsAsFactors=F)

for(i in 1:nrow(rmap_df)) {
  # update non-matching country names within data frame
  tb_df$country[tb_df$country == rmap_df$tb_country[i]] <- rmap_df$rmap_country[i]
}

# now calculate average case counts for each country
tb_means <- tb_df %>% group_by(country) %>% summarise(avg_cases = mean(cases, na.rm = TRUE))

# Sort the averages in descending order
tbm_sorted <- arrange(tb_means, desc(avg_cases))
```



```{r, echo = FALSE, fig.height=6, fig.width=9.4, message=FALSE, warning=FALSE}
# geoplot average case counts + rates of infection, 1995-2013

# par(mar=c(4, 3, 3, 1) + 0.1)

par(mai=c(1.5,2,0,0.42))

# plot geomap for number of cases by country
capture.output(spdf <- joinCountryData2Map(tbm_sorted, joinCode="NAME",
                                           nameJoinColumn="country", verbose=FALSE),
               file='NUL')

mapParams <- mapCountryData(spdf, nameColumnToPlot="avg_cases",
                            catMethod=c(0,1000,5000,10000,20000,50000,100000,250000,500000, 
                                        1000000), addLegend=FALSE, numCats = 9, aspect = 1.3,
                            mapTitle = "")

mapParams$legendText <- c('0-1000',
                          '1001-5000',
                          '5000-10000',
                          '10001-20000',
                          '20001-50000',
                          '50001-100000',
                          '100001-250000',
                          '250001-500000',
                          '500001-1000000')

do.call( addMapLegendBoxes, c( mapParams, horiz=FALSE, title="# TB Cases", 
                               cex = .9, pt.cex = 1.5))
```


## Average Annual TB Infection Rates (1995 - 2013)


```{r, echo = FALSE}
# get tb rates of infection data
tbrate_means <- sqlQuery(con, "SELECT DISTINCT country, AVG(rate) FROM tb_rates GROUP BY country ORDER BY AVG(rate) DESC", stringsAsFactors=F)

# rename second column to meaningful name
colnames(tbrate_means)[2] <- "avg_rate"

tbrate_means$avg_rate <- tbrate_means$avg_rate * 100000
```

```{r, fig.height=6, fig.width=9.4, echo = FALSE, message=FALSE, warning=FALSE}

# plot geomap for tb infection rates by country

# par(mar=c(4, 3, 3, 1) + 0.1)
par(mai=c(1.5,2,0,0.42))

capture.output(spdf <- joinCountryData2Map(tbrate_means, joinCode="NAME",
                                           nameJoinColumn="country", verbose = FALSE),
               file='NUL')

mapParams <- mapCountryData(spdf, nameColumnToPlot="avg_rate",
                            catMethod=c(0,10,50,75,100,150, 
                                        200, 300, 500, 1000), 
                            addLegend=FALSE, numCats = 9, aspect = 1.3,
                            mapTitle = "")

mapParams$legendText <- c('0-10',
                          '11-50',
                          '51-76',
                          '76-100',
                          '101-150',
                          '151-200',
                          '201-300',
                          '301-500',
                          '501 - 1000')

do.call( addMapLegendBoxes, c( mapParams, horiz=FALSE,  title="TB Cases/100K People",
                               cex = .9, pt.cex = 1.5))
```

## Average Annual TB Case Counts (1995 - 2013)

```{r, echo = FALSE, fig.height=6, fig.width=8.4,}

par(mai=c(1.5,2,0.2,0.42))

x <- arrange(head(tbm_sorted, n=20), avg_cases)
Z <- x$avg_cases

barplot(x$avg_cases, names = x$country, las=2, horiz = TRUE, xaxt = "n",
        xlim=c(0,500000), col = rainbow(20), cex.axis = .8, cex.names = .7,
        main = "Top 20 Countries")

axis(1, las=2, axTicks(1), format(axTicks(1), scientific = F))

rm(Z)
rm(x)
```

## Average Annual TB Infection Rates (1995 - 2013)

```{r, echo = FALSE, fig.height=6, fig.width=8.4}

par(mai=c(1.5,2,0.2,0.42))

x <- arrange(head(tbrate_means, n=20), avg_rate)
Z <- x$avg_rate

barplot(x$avg_rate, names = x$country, las=2, horiz = TRUE, xaxt = "n",
        xlim=c(0,400), col = rainbow(20), cex.axis = .8, cex.names = .7,
        main = "Top 20 Countries",
        sub = "Cases per 100,000 People")
axis(1, las=2, axTicks(1), format(axTicks(1), scientific = F))

rm(Z)
rm(x)
```

## Cause For Alarm

```{r, echo = FALSE}

# get tb rates for all countries
tbr_df <- sqlQuery(con, "SELECT * FROM tb_rates", stringsAsFactors=F)

tbr_df$rate <- tbr_df$rate * 100000

# -----------------------------------------------------------------
# Sort the rates in descending order for ALL years
highrates <- arrange(tbr_df, desc(rate), country, year)
```


```{r, echo = FALSE}
rateT20 <- head(tbrate_means, n = 20)
casesT20 <-  head(tbm_sorted, n = 20)

high_cr <- data.frame()

for(i in 1:nrow(rateT20)) {
  # find country names that occur in both top 20 case counts and top 20 infection rates
  high_cr <- rbind(high_cr, 
                   data.frame(casesT20$country[casesT20$country == rateT20$country[i]]))
}

# rename second column to meaningful name
colnames(high_cr)[1] <- "country"

# display VERY HIGH RISK countries
kable(high_cr, row.names = TRUE, caption = "Countries With Both High TB Case Counts & High Infection Rates")
```

## Trends

```{r, echo = FALSE}
par(mai=c(.5,2,0.4,0.42))

y_cases <- tb_df %>% group_by(year) %>% summarise(tot_cases = sum(cases, na.rm = TRUE))

Z <- y_cases$tot_cases

# plot total cases by year, using case count as basis of colorcode for bar
barplot(y_cases$tot_cases, names = c(as.character(y_cases$year)), las=3, 
        ylim=c(0,4000000), col = topo.colors(max(Z))[Z], cex.axis = .8, cex.names = .8,
        main = 'Total # of Annual TB Cases, 100 Countries, 1995 - 2013', ylab = "# Cases" )
rm(Z)
rm(y_cases)
```

__Worldwide TB case counts and infection rates rose sharply during the (2006 - 2008) time period.__

## Evidence of Worldwide TB Case Spike 2006 - 2008
```{r, echo=FALSE, fig.height=6, fig.width=10.4}
# Plotting 2 rows x 3 columns of plots = 6 plots total

# par(mfrow=c(3,2))
    
# par(mfrow = c(3,4), oma = c(2, 2, 0, 0), mar=c(2, 2, 2, 2) + 2)

par(mfrow = c(3,4), oma = c(1, 1, 0, 0), mar=c(1.5, 1.5, 0, 1.5) + 2)

# ----------------------------
# __China__
t1 <- subset(tb_df, country == 'China')

ytop <- max(t1$cases, na.rm = TRUE)

barplot(t1$cases, names = "", las=3,ylim=c(0,ytop), 
        col = 'blue', main = 'China' )

# add axis labels
mtext(side = 1, text = "1995 - 2013", line = 1)
mtext(side = 2, text = "# Cases", line = 2)

# ----------------------------
# __Indonesia__
t1 <- subset(tb_df, country == 'Indonesia')

ytop <- max(t1$cases, na.rm = TRUE)

barplot(t1$cases, names = "", las=3, ylim=c(0,ytop), 
        col = 'blue', main = 'Indonesia')

# add axis labels
mtext(side = 1, text = "1995 - 2013", line = 1)
mtext(side = 2, text = "# Cases", line = 2)

# ----------------------------
# __South Africa__
t1 <- subset(tb_df, country == 'South Africa')

ytop <- max(t1$cases, na.rm = TRUE)

barplot(t1$cases, names = "", las=3, ylim=c(0,ytop), 
        col = 'blue', main = 'South Africa')

# add axis labels
mtext(side = 1, text = "1995 - 2013", line = 1)
mtext(side = 2, text = "# Cases", line = 2)

# ----------------------------
# __Russia__
t1 <- subset(tb_df, country == 'Russia')

ytop <- max(t1$cases, na.rm = TRUE)

barplot(t1$cases, names = "", las=3, ylim=c(0,ytop), 
        col = 'blue', main = 'Russia' )

# add axis labels
mtext(side = 1, text = "1995 - 2013", line = 1)
mtext(side = 2, text = "# Cases", line = 2)

# ----------------------------
# __Brazil__
t1 <- subset(tb_df, country == 'Brazil')

ytop <- max(t1$cases, na.rm = TRUE)

barplot(t1$cases, names = "", las=3, ylim=c(0,ytop), 
        col = 'blue', main = 'Brazil' )

# add axis labels
mtext(side = 1, text = "1995 - 2013", line = 1)
mtext(side = 2, text = "# Cases", line = 2)

# ----------------------------
# __Poland__
t1 <- subset(tb_df, country == 'Poland')

ytop <- max(t1$cases, na.rm = TRUE)

barplot(t1$cases, names = "", las=3, ylim=c(0,ytop),
        col = 'blue', main = 'Poland' )

# add axis labels
mtext(side = 1, text = "1995 - 2013", line = 1)
mtext(side = 2, text = "# Cases", line = 2)


# ----------------------------
# __Argentina__
t1 <- subset(tb_df, country == 'Argentina')

ytop <- max(t1$cases, na.rm = TRUE)

barplot(t1$cases, names = "", las=3, ylim=c(0,ytop), 
        col = 'blue', main = 'Argentina' )

# add axis labels
mtext(side = 1, text = "1995 - 2013", line = 1)
mtext(side = 2, text = "# Cases", line = 2)

# ----------------------------
# __Egypt__
t1 <- subset(tb_df, country == 'Egypt')

ytop <- max(t1$cases, na.rm = TRUE)

barplot(t1$cases, names = "", las=3, ylim=c(0,ytop), 
        col = 'blue', main = 'Egypt' )

# add axis labels
mtext(side = 1, text = "1995 - 2013", line = 1)
mtext(side = 2, text = "# Cases", line = 2)


# ----------------------------
# __USA__
t1 <- subset(tb_df, country == 'USA')

ytop <- max(t1$cases, na.rm = TRUE)

barplot(t1$cases, names = "", las=3, ylim=c(0,ytop), 
        col = 'blue', main = 'USA' )

# add axis labels
mtext(side = 1, text = "1995 - 2013", line = 1)
mtext(side = 2, text = "# Cases", line = 2)

# ----------------------------
# __Mali__
t1 <- subset(tb_df, country == 'Mali')

ytop <- max(t1$cases, na.rm = TRUE)

barplot(t1$cases, names = "", las=3, ylim=c(0,ytop), 
        col = 'blue', main = 'Mali' )

# add axis labels
mtext(side = 1, text = "1995 - 2013", line = 1)
mtext(side = 2, text = "# Cases", line = 2)

# ----------------------------
# __Guatemala__
t1 <- subset(tb_df, country == 'Guatemala')

ytop <- max(t1$cases, na.rm = TRUE)

barplot(t1$cases, names = "", las=3, ylim=c(0,ytop), 
        col = 'blue', main = 'Guatemala')

# add axis labels
mtext(side = 1, text = "1995 - 2013", line = 1)
mtext(side = 2, text = "# Cases", line = 2)

rm(t1)
rm(ytop)
par(mfrow=c(1,1))
```

## Per Capita Metrics vs. TB Infection Rates

Linear least squares regression finds little relation between per capita metrics and TB infection rates

```{r, echo = FALSE}
# get life expectancy data set
lifexp_df <- sqlQuery(con, "SELECT * FROM life_exp", stringsAsFactors=F)
```

```{r, echo = FALSE, message=FALSE, warning=FALSE}
# create a copy of tbr_df data frame
t.df <- tbr_df

# add life expectancy to tb_rates dataframe
t.df$life_exp <- lifexp_df$life_exp
```

```{r, echo= FALSE}
# get healthcare data set
hc_df <- sqlQuery(con, "SELECT * FROM percap_hc", stringsAsFactors=F)

# add healthcare data to tb_rates dataframe
t.df$hc <- hc_df$percap_hc
```

```{r, echo = FALSE}
# get gni data set
gni_df <- sqlQuery(con, "SELECT * FROM percap_gni", stringsAsFactors=F)

# add gni data to tb_rates dataframe
t.df$gni <- gni_df$percap_gni
```

```{r, echo = FALSE}
# get electricity data set
eacc_df <- sqlQuery(con, "SELECT * FROM perc_e_acc", stringsAsFactors=F)

# fetch TB rates data for 3 yrs only: 2000, 2010, 2012
t2.df <- sqlQuery(con, "SELECT * FROM tb_rates WHERE year IN(2000, 2010, 2012)")
t2.df$rate <- t2.df$rate * 100000

# add electricity data to tb_rates dataframe
t2.df$elec <- eacc_df$perc_e_acc
```

```{r, echo = FALSE}
# get avg yrs schooling data set
school_df <- sqlQuery(con, "SELECT * FROM yrs_school", stringsAsFactors=F)

# fetch TB rates data for only yrs: 2000, 2005 - 2012
t3.df <- sqlQuery(con, "SELECT * FROM tb_rates WHERE year 
                 IN(2000, 2005, 2006, 2007, 2008, 2009, 2010, 2011, 2012)")

t3.df$rate = t3.df$rate * 100000

# add school data to tb_rates dataframe
t3.df$school <- school_df$yrs_school
```


```{r, echo = FALSE, warning=FALSE, message= FALSE}
# par(mfrow = c(2,3), oma = c(2, 2, 0, 0), mar=c(1.5, 1.5, 1, 1) + 1)

require(gridExtra)


# plot using ggplot
p1 <- ggplot(t.df, aes(x = life_exp, y = rate)) + 
  geom_point() +
  stat_smooth(method = "lm", col = "red") +
  xlab("Life Expectancy") +
  ylab("TB Infection Rate / 100K")

# plot using ggplot
p2 <- ggplot(t.df, aes(x = hc, y = rate)) + 
  geom_point() +
  stat_smooth(method = "lm", col = "red") +
  xlab("Per Cap. Healthcare Spend") +
  ylab("TB Infection Rate / 100K")
  
# plot using ggplot
p3 <- ggplot(t.df, aes(x = gni, y = rate)) + 
  geom_point() +
  stat_smooth(method = "lm", col = "red") +
  xlab("Per Cap. GNI") +
  ylab("TB Infection Rate / 100K")

# plot using ggplot
p4 <- ggplot(t2.df, aes(x = elec, y = rate)) + 
  geom_point() +
  stat_smooth(method = "lm", col = "red") +
  xlab("Pop. % w/ Electricity") +
  ylab("TB Infection Rate / 100K People")

# plot using ggplot
p5 <- ggplot(t3.df, aes(x = school, y = rate)) + 
  geom_point() +
  stat_smooth(method = "lm", col = "red") +
  xlab("Avg Years of Schooling") +
  ylab("TB Infection Rate / 100K People")  

grid.arrange(p1, p2, p3, p4, p5, ncol=3, nrow = 2)
```

## Per Capita Metrics vs. TB Infection Rates

However, there clearly are __*major*__ differences in the per capita metrics of countries with low and high TB infection rates:

| Type of Country    | Life Exp. | HC Exp. |   GNI   | Elec. Acc.| Schooling |
| ------------------ | --------- | ------- | ------- | --------- | --------- |
| Low TB Rate        | 76.63 yrs | $966.90 | $25,640 |  100.00%  |  9.05 yrs |
| High TB Rate       | 59.66 yrs | $114.20 | $ 3,280 |   37.45%  |  6.07 yrs |
| Difference:        | 16.97 yrs | $852.70 | $22,360 |   62.55%  |  2.98 yrs |

<br> 

__*So what is going on here?*__

- A relationship likely __*DOES*__ exist between the per capita metrics and TB infection rates.

- However, if such a relationship exists, it must be __*non-linear*__.

## Conclusions

- Linear least squares modeling is __*not*__ useful for identifying relationships between per capita metrics and TB infection rates.

- However, the differences in per capita metrics of countries with low and high TB infection rates indicates that some sort of relationship likely __*DOES*__ exist.

- Therefore, other types non-linear modeling would be required 
  
<br> 
  
__*Access the Full Report & Separate Data Acquisition Module at*__:

- Analysis & Writeup: https://rpubs.com/jt_rpubs/178917

- Data Acquisition Module: https://rpubs.com/jt_rpubs/178913


```{r, echo = FALSE}
# close the database connection
odbcClose(con)
```

